Transitioning towards Continuous Delivery in the B2B Domain: A Case Study

### 2014-12-07 paper.tex
Added content

\begin{abstract}        % give a summary of your study
Delivering value to customers in real-time requires a company to utilize real-time deployment of software to expose features to users faster, and to shorten the feedback loop. This allows for faster reaction times, ensuring the development is focused on features providing real value. Continuous delivery is a development practice where the software functionality is deployed continuously to customer environment. Although this practice has been established in some domains such as B2C mobile software, the B2B domain imposes specific challenges. This article imposes a case study that is conducted in a medium-sized software company in the B2B domain. The results suggest that technical challenges are only one part of the challenges a company encounters in this transition. For continuous delivery, the company must also address challenges related to the customer and procedures. The core challenges are caused by having multiple customers with diverse environments and unique properties, whose business depends on the software product. Some customers also require to perform manual acceptance testing, which slows down production deployments. The benefits found from continuous delivery support the case company in solving many of its business problems. The company can expose the software functionality to the customers from an earlier stage, and better utilize customer feedback.
%                         please supply keywords within your abstract
\keywords {Continuous Delivery, Continuous Deployment, Development Process, B2B, Case Study}
\end{abstract}
%
\section{Introduction}

@@

Continuous delivery is a design practice aiming to shorten the delivery cycles and by automating the deployment process. It is an extension to continuous integration, where the software functionality is deployed frequently to customer environment, and the software is developed in a way that it is always ready for release. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, continuous delivery adds automated acceptance testing and automated deployment.

@@

\subsection{Background and Related Work}
Continuous delivery is an extension to continuous integration, where the software functionality is deployed frequently at customer environment. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, often multiple times a day, continuous delivery adds automated acceptance testing and deployment to a staging environment. The purpose of continuous delivery is that as the deployment process is automated, it reduces human error, documents required for the build and increases confidence that the build works \cite{cdbook}. Continuous delivery solves the problem of how to deliver an idea to users as quickly as possible.

@@

%\begin{figure}
% use the \includegraphics command from the graphics package
% (N.B. put \usepackage{graphics} into your preamble then)
% for inclusion of your images. A sample call could read:
% \includegraphics{myimage}
%\vspace{2.5cm}
%\caption{This is the caption of the figure displaying a white eagle and
%a white horse on a snow field}
%\end{figure}
%delivery vs deployment
Continuous delivery differs from continuous deployment. Both continuous delivery and continuous deployment include automated deployment to a staging environment. Continuous deployment includes deployment to a production environment, while in continuous delivery the deployment to a production environment is done manually. The purpose of continuous delivery is to prove that every build is deployable \cite{cdbook}. While it necessarily doesn't mean that teams release often, keeping the software in a state where a release can be made instantly is often seen beneficial.

%Barriers in CI->CD transition: stairway
Holmström Olsson et al. research the transition phase from continuous integration to continuous delivery, and identify barriers that companies need to overcome to achieve the transition \cite{olsson2012climbing}. One such barrier is the custom configuration at customer sites. Maintaining customized solutions and local configurations alongside the standard configurations creates issues. The second barrier is the internal verification loop, that has to be shortened not only to develop features faster but also to deploy fast. Finally, the lack of transparency and getting an overview of the status of development projects is seen as a barrier.

@@

The purpose of this study is to seek ways to improve the product development process of two software products of the case company in the B2B domain, utilizing a practice known as continuous delivery. This includes analysing the challenges and benefits of this practice. Existing documented implementations of continuous delivery are primarily executed in the B2C domain, often with a Software as a Service (SaaS) product. The focus of this study is in the B2B domain, with two software products that are not used as SaaS and are developed directly to other companies acting as customers. The characteristics of this domain are typically longer release cycles and customer owned environments.

@@

\subsection{RQ1}

@@

\subsubsection{Technical challenges}
The technical challenges for continuous delivery are derived from the interviews. A part of the interview focuses on the current deployment process and customer interaction of the case company. The current deployment process, customer interaction and challenges related to them are then used as a basis for analysing challenges that influence continuous delivery.

%\begin{center}
%\begin{table}[htb]
%    \begin{tabular}{ | p{12cm} |}
%    \hline
%    \textbf{Specific problem} \\ \hline
%    Downtime is critical for certain customers \\ \hline
%    Automated testing has to be built on top of a matured software product \\ \hline
%    Software is often integrated to multiple third party applications \\ \hline
%    Software is often accompanied by multiple external components \\ \hline
%    There exists multiple different configurations due to having multiple customers with different specifications \\ \hline
%    Transferring the software product to diverse customer-owned environments requires different deployment configurations \\ 
%    \hline
%    \end{tabular}
%    \caption{Technical challenges in continuous delivery.}
%    \end{table}
%\end{center}
%Downtime
Downtime of the case company's products can be fatal. According to the Dialog product owner, downtime causes end-users being unable to perform their job. Downtime can also interrupt ongoing customer tasks, possibly losing critical data in the progress. With CDM, downtime can cause other integrated systems to shut down due to too many failing requests. Currently the deployment time for both projects is negotiated with the customer to prevent these cases, and the version deployments are done when the system can be closed for a short period of time. In the continuous delivery process the deployment pipeline has to be designed to address the issue of downtime.

%-testing costs could be pushed at least partially to users
The developers perceive automated testing and test environments to be the largest technical task. Since both of the case company's software products have been in development for a long time, building a viable test automation has many challenges with architectural decisions and available workload. The developers state that building a sufficient test automation is a very laborious process especially due to the maturity of the software, and are concerned with the maintainability of the test suite. The management is not sure what to test with automatic acceptance testing to validate that a version is valid. Due to the current amount of time spent on manual testing, having an automated test suite is seen as very beneficial.% Humble et al. have also found that the production line should be built along with the application it assembles \cite{humble2006deployment}.

%While this task can, depending on the project and current state of tests, take up most of the resources in transition to continuous delivery, in B2B context some of the testing costs can be pushed partially to customer projects. 
%Automate testing

%The last principle is "Evolve your production line along with the application it assembles". Humble et al. state that attempting to build a full production line before writing any code doesn't deliver any value, so the production line should be built and modified as the application evolves.

%Third party application integration
Both of the case company's software products are integrated to various third party applications and APIs. As CDM is a component that integrates various data sources together, changes to the APIs communicating with these applications must be planned and discussed in advance. Based on the interview results, automatically updating the integrations requires an unduly amount of work considering the results.

@@

\subsubsection{Procedural challenges}
The procedural challenges are analyzed based on the development process documentations of the company and the interviews. In the interviews, a section is dedicated to the current development process of the case company. The development process and its challenges are then used to analyse and identify challenges that influence continuous delivery.

%\begin{center}
%\begin{table}[htb]
%    \begin{tabular}{ | p{12cm} |}
%    \hline
%    \textbf{Specific problem} \\ \hline
%    User acceptance testing environment is a requisite for production release \\ \hline
%    The development process drifts towards small feature branches from long-lived feature branches \\ \hline
%  Triggering the compilation and deployment of a modular project to maintain integrity is hard \\ \hline
%  The software has to be deployed to multiple customers \\ \hline
%  Versioning is affected by having different customer profiles of the product \\ \hline
%  Responsibility of deploying moves towards developers \\ \hline
%  Management and sales loses track of versions \\
%    \hline
%    \end{tabular}
%    \caption{Procedural challenges in continuous delivery.}
%    \end{table}
%\end{center}

@@

\subsubsection{Customer challenges}
The customer challenges are analyzed based on two sections of the interview: customer interaction and the deployment process.

%\begin{center}
%\begin{table}[htb]
%    \begin{tabular}{ | p{12cm} |}
%    \hline
%    \textbf{Specific problem} \\ \hline
%    Some customers are reluctant towards new versions \\ \hline
%    Customers are trained to use a certain version, and modifications confuse the users \\ \hline
%    Changelogs are especially important, since as versions are released faster the customers become less aware on what has changed  \\ \hline
%    Pilot customer is required for developing the continuous delivery process \\ \hline
%    Acceptance testing is performed by both the company and the customers, and requires a lot of resources from the customers \\ \hline
%  Production deployment schedule has to be negotiated with the customer \\ \hline
%  Ongoing critical tasks by users cannot be interrupted by downtime \\ 
%    \hline
%    \end{tabular}
%    \caption{Customer challenges in continuous delivery.}
%    \end{table}
%\end{center}

@@

\subsection{RQ2}

@@

>>>>

\begin{abstract}        % give a summary of your study
Delivering value to customers in real-time requires companies to utilize real-time deployment of software to expose features to users faster, and to shorten the feedback loop. This allows for faster reaction times, ensuring the development is focused on features providing real value. Continuous delivery is a development practice where the software functionality is deployed continuously to customer environment. Although this practice has been established in some domains such as B2C mobile software, the B2B domain imposes specific challenges. This article presents a case study that is conducted in a medium-sized software company in the B2B domain. The results suggest that technical challenges are only one part of the challenges a company encounters in this transition. The company must also address challenges related to the customer and procedures. The core challenges are caused by having multiple customers with diverse environments and unique properties, whose business depends on the software product. Some customers also require to perform manual acceptance testing, which slows down production deployments. The benefits found from continuous delivery support the case company in solving many of its business problems. The company can expose the software functionality to the customers from an earlier stage, and better utilize customer feedback.
%                         please supply keywords within your abstract
\keywords {Continuous Delivery, Continuous Deployment, Development Process, B2B, Case Study}
\end{abstract}
%
\section{Introduction}

@@

Continuous delivery is a design practice aiming to shorten the delivery cycles and by automating the deployment process. It is an extension to continuous integration, where the software functionality is deployed frequently to customer environment. In continuous delivery, the software is developed in a way that it is always ready for release. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, continuous delivery adds automated acceptance testing and automated deployment.

@@

\subsection{Background and Related Work}
Continuous delivery is an extension to continuous integration, where the software functionality is deployed frequently at customer environment. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, often multiple times a day, continuous delivery adds automated acceptance testing and deployment to a staging environment. The purpose of continuous delivery is that as the deployment process is automated, it reduces human error, documents required for the build and increases confidence that the build works \cite{cdbook}. Continuous delivery aims to solve the problem of how to deliver an idea to users as quickly as possible.

@@

%\begin{figure}
% use the \includegraphics command from the graphics package
% (N.B. put \usepackage{graphics} into your preamble then)
% for inclusion of your images. A sample call could read:
% \includegraphics{myimage}
%\vspace{2.5cm}
%\caption{This is the caption of the figure displaying a white eagle and
%a white horse on a snow field}
%\end{figure}
%delivery vs deployment
Continuous delivery differs from continuous deployment. Both continuous delivery and continuous deployment include automatic deployment to a staging environment. Continuous deployment includes deployment to a production environment, while in continuous delivery the deployment to a production environment is done manually. The purpose of continuous delivery is to prove that every build is deployable \cite{cdbook}. While it necessarily doesn't mean that teams release often, keeping the software in a state where a release can be made instantly is often seen beneficial.

%Barriers in CI->CD transition: stairway
Olsson et al. investigate the transition phase from continuous integration to continuous delivery, and identify barriers that companies need to overcome to achieve the transition \cite{olsson2012climbing}. One such barrier is the custom configuration at customer sites. Maintaining customized solutions and local configurations alongside the standard configurations creates issues. The second barrier is the internal verification loop, that has to be shortened not only to develop features faster but also to deploy fast. Finally, the lack of transparency and getting an overview of the status of development projects is seen as a barrier.

@@

%The purpose of this study is to seek ways to improve the product development process of two software products of the case company in the B2B domain by utilizing continuous delivery. This includes analysing the challenges and benefits of this practice. Existing documented implementations of continuous delivery are primarily executed in the B2C domain, often with a Software as a Service (SaaS) product. The focus of this study is in the B2B domain, with two software products that are not used as SaaS and are developed directly to other companies acting as customers. The characteristics of this domain are typically longer release cycles and customer owned environments.

@@

@@

\subsection{Technical challenges}
The technical challenges for continuous delivery are derived from the interviews. A part of the interview focuses on the current deployment process and customer interaction of the case company. The current deployment process, customer interaction and challenges related to them are then used as a basis for analysing challenges that influence continuous delivery.

\begin{table}[htb]
    \begin{tabular}{ | p{12cm} |}
    \hline
    \textbf{Specific problem} \\ \hline
    Downtime is critical for certain customers \\ \hline
    Automated testing has to be built on top of a matured software product \\ \hline
    Software is often integrated to multiple third party applications \\ \hline
    Software is often accompanied by multiple external components \\ \hline
    There exists multiple different configurations due to having multiple customers with different specifications \\ \hline
    Transferring the software product to diverse customer-owned environments requires different deployment configurations \\ 
    \hline
    \end{tabular}
    \caption{Technical challenges in continuous delivery.}
    \end{table}

Downtime of the case company's products can be fatal. According to the Dialog product owner, downtime causes end-users being unable to perform their job. Downtime can also interrupt ongoing customer tasks, possibly losing critical data in the progress. Currently the deployment time for both projects is negotiated with the customer to prevent these cases, and the version deployments are done when the system can be closed for a short period of time.

%-testing costs could be pushed at least partially to users
The developers perceive automated testing and test environments to be the largest technical task. Building a viable test automation has many challenges with architectural decisions and available workload. The developers state that building a sufficient test automation is a very laborious process especially due to the maturity of the software, and are concerned with the maintainability of the test suite. The management is not sure what to test with automatic acceptance testing to validate that a version is valid.

Both of the case company's software products are integrated to various third party applications and APIs. Changes to the interfaces communicating with these applications must be planned and discussed in advance. Based on the interview results, automatically updating the integrations requires an unduly amount of work considering the results.

@@

\subsection{Procedural challenges}
The procedural challenges are analyzed based on the development process documentations of the company and the interviews. In the interviews, a section is dedicated to the current development process of the case company. The development process and its challenges are then used to analyse and identify challenges that influence continuous delivery.

\begin{table}[htb]
    \begin{tabular}{ | p{12cm} |}
    \hline
    \textbf{Specific problem} \\ \hline
    User acceptance testing environment is a requisite for production release \\ \hline
    The development process drifts towards small feature branches from long-lived feature branches \\ \hline
  Triggering the compilation and deployment of a modular project to maintain integrity is hard \\ \hline
  The software has to be deployed to multiple customers \\ \hline
  Versioning is affected by having different customer profiles of the product \\ \hline
  Responsibility of deploying moves towards developers \\ \hline
  Management and sales loses track of versions \\
    \hline
    \end{tabular}
    \caption{Procedural challenges in continuous delivery.}
    \end{table}

@@

\subsection{Customer challenges}
The customer challenges are analyzed based on two sections of the interview: customer interaction and the deployment process.

\begin{table}[htb]
    \begin{tabular}{ | p{12cm} |}
    \hline
    \textbf{Specific problem} \\ \hline
    Some customers are reluctant towards new versions \\ \hline
    Customers are trained to use a certain version, and modifications confuse the users \\ \hline
    Changelogs are especially important, since as versions are released faster the customers become less aware on what has changed  \\ \hline
    Pilot customer is required for developing the continuous delivery process \\ \hline
    Acceptance testing is performed by both the company and the customers, and requires a lot of resources from the customers \\ \hline
  Production deployment schedule has to be negotiated with the customer \\ \hline
  Ongoing critical tasks by users cannot be interrupted by downtime \\ 
    \hline
    \end{tabular}
    \caption{Customer challenges in continuous delivery.}
    \end{table}

@@

\subsection{Benefits to the case company}

This section addresses the RQ2 (i.e. How do continuous delivery and continuous experimentation benefit the case company?). A major problem found in the interviews is that currently the reliability towards new versions is low. The low reliability both increases customers reluctancy towards version updates, and increases the amount of user access testing that is performed after version release. This makes the duration of building a new version relatively long and unstable, since the bug amounts differ upon each version deployment. The version releases are also often performed in a hurry, which also increases the error rate. The low reliability is caused by multiple reasons: features are occasionally forgotten and not tested, manual builds introduce human error factor, the knowledge of manual steps required for a succesful build is not shared and compiling modular products is hard.

%Features are forgotten
Versions are occasionally forgotten from the UAT phase due to the lack of comprehensive automated testing and the broad scale of features in both software products. These features can then remain broken or contain bugs when the users start using the new version. Even though these features get fixed relatively fast, the introduced bugs both increase the customer reluctancy towards new versions, and decrease the reliability of both the team and the customer. This is fundamentally caused by the lack of quality assurance before the release. Adopting a test automation solves this issue, as long as tests are written for every feature.

%Human error factor
The manual build process involves human error factor, which in the case company is the primary cause for the majority of bugs in new versions. One example of such an error is using the wrong commands to compile and package the project, which causes wrong binaries getting into the new version. Combined with the lack of comprehensive automated testing, each deployment is a new error-prone experiment. Important factors increasing the rate of human error are partially missing documentation and the complexity of the build process. According to a developer:
\begin{aquote}{CDM developer}
"There are many components that have to be configured manually. The documents for how to execute these manual actions exist, but many things still rely on memory."
\end{aquote}

%Some steps are memoralized
While some steps of the deployment process are documented, there are still parts that are memorized by developers. Since some of the manual steps are only memorized by certain developers, the deployment requires assistance from these persons. With continuous delivery only a handful of developers might have knowledge of the entire build deployment configuration, but everyone are able to trigger the deployment process. Automating the entire deployment process also reduces the human error factor and necessity for documentation, since deploying can be performed at a press of a button.

%-Compiling a modular product is hard
%-Some wrong binaries might be compiled under the new version, have to manually check these
The interviewees perceive compiling and building CDM a tricky task. CDM is a modular product compiled out of many individual packages, which each have a unique version. Whenever a developer updates a version, attention has to be paid that the newest version of the package is present in the compiled application. By manually triggering the build progress, different flags can be used to define if a package is loaded from the artifactory or local environment. If the versions of these two environments differ, the product might get built with an incoherent composition of packages. With continuous delivery a good CI tool will handle the dependency management \cite{cdbook}.

%-Many customer environments with different configurations

%nämä haasteita:
%    Software is often integrated to multiple third party applications   
%    Software is often accompanied by multiple external components   
%    There exists multiple different configurations due to having multiple customers with different specifications   
%    Transferring the software product to diverse customer-owned environments requires different deployment configurations

%-Allows for faster feedback collection
%Haaste: nykyään joudutaan odottamaan kauan palautetta, kun versioita viedään niin vähän. Myöskään testeistä ei saada nopeasti palautetta. Devaajia motivoisi enemmän jos koodin näkisi menevän käyttöön päivittäin.
With the current deployment model, release cycle length ranges from two to four weeks. The feedback cycle is therefore quite long, and non-critical bugs reported by customers might take until the next release to fix. By adopting continuous delivery, feedback can be received a lot faster from the customers. Implementing the automated tests also adds a layer of feedback. The developers state that they would be more motivated if code could be deployed faster. Deploying the code makes the developers feel like they have achieved something concrete. However, developing the software so that it is always ready for release also improves the developers confidence towards the created code and product.

%-Requires less time to deploy
Automating the deployment also reduces the time required for deploying a version. The product owners state that building and deploying a version can take up to a day, which is seen as way too long. The primary reason taking up the time is caused by having to manually compile, test and bugfix the versions. With continuous delivery, the amount of manual work can be diminished to couple of button clicks defining which customer profile to deploy. Continuous delivery therefore allows the release of changes in time spans of hours, also allowing the most important features and fixes to be delivered faster. With faster problem fixing due to increases in bug discover and fixing rate, the software eventually ends up with fewer bugs. %The primary reason for

%Duration
Along with the deployment, environment installations take relatively long with manual installation of each component. With continuous delivery, starting the application in a new environment is ideally just generating configuration information describing the environment's unique properties, and starting the automated deployment process to both prepare the new environment and to deploy the correct version \cite{cdbook}. This allows for much more flexibility of deployments, where versions can be deployed when it is convenient to do so.

%-Downtime can be eliminated %Canary release
The prolonged deployment time also causes more downtime. Downtime is not currently perceived as an issue in itself, as it is negotiated with the customer and happens during the hours where downtime is acceptable. With continuous delivery, the company gains more incentive to reduce or eliminate downtime. In the case of CDM, downtime can be eliminated by utilizing for example parallel environments or canary releases \cite{cdbook}. With canary releases, a new version is rolled to a subset of the production server, and a part of the users are routed to the new version. If the new version is then deemed to work as intended, all of the users are routed to the new version.

%issues tracking status
One of the issues is that customers are having problems tracking the status of the development process in projects. This is caused by not deploying features that are not complete, and the relatively long length of the software delivery cycles. The customer is better kept on track of the development with continuous delivery, as the process is more transparent. In projects with a set duration it is much easier for the customer to understand the status of features when they are deployed in constant intervals.

%As the customer is constantly updated with newer versions, in projects with a set duration i

%Ongelma: uuden ympäristön asennus kestää erittäin kauan (useita työpäiviä)
%It should be a simple task to start your application in a new environment—ideally
%just a matter of commissioning the machines or virtual images and creating some
%configuration information that describes the environment’s unique properties.
%Then you should be able to use your automated deployment process to prepare
%the new environment for deployment and deploy the chosen version of your
%application to it. \cite{cdbook}

To sum up the findings, continuous delivery provides an answer to many of the challenges faced in the current deployment process and product development in general. The management considers improving the deployment process to be one of the most important improvements. According to the findings, continuous delivery mainly increases the speed, quality, and capacity of the development. Speed and capacity are ensured by faster delivery, while quality is increased by the automated testing and faster feedback. Smaller problems can be quickly fixed without spending unnecessary time on manually deploying a new version to the customer, and bigger changes only take as long as the implementation requires. After the initial investion, the practice will eventually allow the company to spend less money on management and operations, because both meeting costs and unnecessary repetitive work can be eliminated. %D

@@

The study also suggests that continuous delivery corresponds to many of the case company's primary needs. The issues related to the deployment process are considered very important by both of the teams. The issues include low reliability in new versions, human error factors when performing version releases and deployments, and long feedback cycles. One of the main benefits of continuous delivery is that the software is kept in a state where it is always ready for deployment \cite{cdbook}, even if the actual deployments are not made more often.

%To theory
%To practice
The issues into which the benefits are mapped were found by researching the current deployment process and challenges faced in the development process. An unexpectedly large part of the major issues stated by the interviewees are related to deploying the software, and it was identified as one of the major challenges in the current development. The benefits found from continuous delivery, which were sought from existing literature \cite{cdbook, neely2013continuous, humble2006deployment}, matched the challenges surprisingly well. By taking the time to map the actual encountered problems to solutions, the concept of continuous delivery can be better sold to the management of companies.
